\documentclass[11pt]{article}

\usepackage{url}
\setlength{\parskip}{0.5cm plus4mm minus3mm}

\setlength{\parindent}{0cm}

\textwidth=6.4in
\textheight=8.5in
\hoffset=-0.7in
\voffset=-0.7in

\title{E4D tutorial}
\author{Alain Plattner}


\begin{document}

\maketitle

\section{Introduction}
E4D is a powerful electrical resistivity tomography program written by
Tim Johnson at PNNL. It can be freely downloaded and installed from
\url{https://e4d.pnnl.gov/}. The program's strength are that it is
parallelized and can therefore solve large problems (many inversion
cells) very quickly. The disadvantage of E4D is that it is harder to
learn than for example BERT.

E4D uses several types of configuration files which we will discuss in the
following. The Matlab/Octave scripts that come with this tutorial will
be helpful to generate the content for many of these input files.

In this tutorial we will go step by step, starting with generating a
simple mesh and moving through calculating inversions. Preparing the
input files for complex meshes is the hardest part of using E4D, so we
will go step by step from simple (but usually useless) meshes to
complicated meshes.

Generally the work flow for using E4D will be to first construct a
mesh, and then run the inversion on a prepared mesh. E4D always
requires an input file named \verb+e4d.inp+. In this input file, the
user specifies what needs to be done (e.g. mesh generation,
inversion), and the names of the configuration and data files.


\section{Mesh generation}\label{meshsec}
On your desktop, or in any folder you would like to work in, create a
file named \verb+e4d.inp+. You can do that for example with emacs by
running in your folder

\qquad \verb+emacs e4d.inp+ 


\subsection{My first mesh}\label{simplemesh}

To make e4d create a mesh, the first line of \verb+e4d.inp+ needs to
be the number 1. The second line needs to be the name of our mesh
configuration file. Let's call it \verb+myfirstmesh.cfg+. So make your
\verb+e4d.inp+ text file look like

\verb+1+\\
\verb+myfirstmesh.cfg+

Of course, we now need to create the mesh configuration file
\verb+myfirstmesh.cfg+. To do that, create it again using emacs

\qquad \verb+emacs myfirstmesh.cfg+

This file will now be a bit more complicated than the input file.
A mesh file will always consist of the following components:
\begin{itemize}
\item The mesh quality (how deformed can the tetrahedra be), and
  maximum tetrahedra volume
\item the bottom elevation of the mesh
\item some information on how to build the mesh (this will always be
  the same so don't worry about this now)
\item The points around which the mesh will be generated including
  electrode positions, topography, boundary points for the internal
  and external zones, electrode depth points for mesh refinement
  around the electrodes, etc.
\item Internal planes to delineate between mesh zones
\item holes (I have not yet worked with holes in the mesh so we will
  skip this in this tutorial)
\item zone descriptions
\item information about writing a paraview/exodus file
\end{itemize}

Let's start with the simplest possible configuration. We first need to
learn how we write points. A point always has the form

\emph{pointnumber xcoord ycoord zcoord type}

\emph{type} can be subsurface point (0), or surface point (1) or
external boundary point (2).

The external boundary points define the outside boundary of our region
so we will always need them.

Let's write our \verb+myfirstmesh.cfg+ file. Let's make the mesh
quality 1.3 and the maximum volume $10^{12}$. The surface will be
defined by the external boundary points (which by definition are on
the surface), and the bottom will be at -20 m. We want the external
boundary to be at at four points (-10/-30/0), (50/-5/1), (40/70/-1.2),
and (-20/60/0). The units of these coordinates are meters.

\newpage
So the lines in our file \verb+myfirstmesh.cfg+ will be:

\verb+1.3 1e12   (mesh quality, max volume)+\\
\verb++\\
\verb+-20   (bottom elevation of mesh)+\\
\verb+1+\\
\verb+"tetgen"+\\
\verb+"triangle"+\\
\verb++\\
\verb+4   (number of points)+\\
\verb+1 -10 -30 0 2   (first boundary point)+\\
\verb+2 50 -5 1 2  (second boundary point)+\\
\verb+3 40 70 -1.2 2  (etc)+\\
\verb+4 -20 60 0 2+\\
\verb++\\
\verb+0   (no internal planes)+\\
\verb++\\
\verb+0   (no holes)+\\
\verb++\\
\verb+1   (one zone)+\\
\verb+1 0 0 -2 50 0.01   (point in zone, max mesh size this zone, conductivity)+\\
\verb++\\
\verb+1   (save mesh as exodus file)+\\
\verb+"bx"  (use the program "bx")+\\
\verb++\\
\verb+1   (mesh translation option)+


\vspace{2cm}

If you want you can omit the comments in parenthesis. These will not
be used by the program but it may make it easier for you to edit the
mesh configuration file.

In this file we needed to define parameters such as maximum tetrahedra
volume for each zone (at this point we only have one zone), and the
conductivity. To do this we needed to pick a random point within our
zone, we picked (0/0/-2), and set the maximum tetrahedra volume, we
selected 50 m$^3$, and for the conductivity, we selected 0.01
S/m. These values are for zone 1, therefore the line read

\emph{1 0 0 -2 50 0.01}

To create the mesh, run in your command line

\qquad \verb+e4d+

The program will automatically look in the input file \verb#e4d.inp#
and use the mesh configuration file you specified there and will
create a lot of auxiliary files. Open the resulting file
\verb#myfirstmesh.exo# using paraview.


\textbf{Exercise:} Play around with the maximum mesh size for the zone
(which we had originally set to 5) and see if you can create a finer
mesh.


\subsection{Adding electrodes}

The mesh we created in Section~\ref{simplemesh} looks great but will
be of limited use when we want to start running inversions. The
minimal thing that we need to do is to add nodes where we have
electrodes.

Let's assume the simple case where we have only 3 electrodes, all on
the surface, at the points (0/0/0), (0/1/0.5), and (0/2/0.2). We will
add these points to the mesh simply by including them in our list of
points. Since our electrodes are all on the surface, their \emph{type}
has the value 1. Our new list of points therefore looks like

\verb+7   (number of points)+\\
\verb+1 -10 -30 0 2   (first boundary point)+\\
\verb+2 50 -5 1 2  (second boundary point)+\\
\verb+3 40 70 -1.2 2  (etc)+\\
\verb+4 -20 60 0 2+\\
\verb++\\
\verb+5 0 0 0 1   (first electrode)+\\
\verb+6 0 1 0.5 1  (second electrode)+\\
\verb+7 0 2 0.2 1  (third electrode)+

The rest of the file \verb+myfirstmesh.cfg+ stays the same as
before. You do not have to insert an empty line between the external
boundary points and the electrodes, but it may make it easier when you
have many electrodes. 

Run in your command prompt

\qquad \verb+e4d+

and open the resulting mesh again with paraview. You will notice that
the mesh looks a bit different. There are smaller tetrahedra where the
electrodes are.

When running electrical resistivity inversions, you will likely run
into problems when the mesh is too coarse around your electrodes. To
create a fine mesh around the electrodes, we will add depth points
underneath the electrodes. These depth points will have the same x and
y coordinates as the electrodes but are a few cm (or m) below the
electrodes (the depth will depend on how fine you will want the mesh
around your electrodes). Let's pick 1 cm in this example (our
electrodes are only 1 m apart). Because
these points are below the surface, their \emph{type} is 0. So your
new list of points will be

\verb+10   (number of points)+\\
\verb+1 -10 -30 0 2   (first boundary point)+\\
\verb+2 50 -5 1 2  (second boundary point)+\\
\verb+3 40 70 -1.2 2  (etc)+\\
\verb+4 -20 60 0 2+\\
\verb++\\
\verb+5 0 0 0 1   (first electrode)+\\
\verb+6 0 1 0.5 1  (second electrode)+\\
\verb+7 0 2 0.2 1  (third electrode)+\\
\verb++\\
\verb+8 0 0 -0.01 0   (first electrode depth point)+\\
\verb+9 0 1 0.49 0  (second electrode depth point)+\\
\verb+10 0 2 0.19 0  (third electrode depth point)+\\

When you run

\qquad \verb+e4d+

with this new \verb+myfirstmesh.cfg+ file, and you look at the result
using paraview, you will see that the mesh around the electrodes is
very fine.

Congratulations, you created your first mesh that could be used for
electrical resistivity tomography. However, to get good results with
reasonable computing power, there is more to learn.

\subsection{Internal boundaries}

One of the problems of electrical resistivity tomography is that our
mesh ends somewhere. This of course is not the case in nature. The
world does not just end. This boundary is therefore artificial.

To avoid problems with this artificial boundary, we will usually have
the external boundary far away, by placing the external boundary
points (the \emph{type} 2 points) far away.

But this leads to the problem that we either can only have a coarse
grid away from the electrodes, or we need to fill the entire volume
with a fine mesh.

Luckily, E4D allows us to create an internal zone with fine mesh
(where we have the electrodes and good coverage), and an external
zone with coarse mesh (which allows us to push the external boundary
far away). In fact, we can have as many zones as we want. We could
make a very fine mesh zone very close to the electrodes, then a
coarser mesh zone a bit further away, and finally a very course mesh
zone surrounding everything.

In this example we will only create two zones, an external zone and an
internal zone, but you can use what you learned here to create more
complex cases.

To create zones we need to create zone boundaries and to create zone
boundaries we need to first pick zone boundary points.

Let's say we want the inner-zone boundary points (-2/-2/-0.3),
(-2/4/0.2), (2/4/0), and (2/-2/0.1). \textbf{Note that these points
  are in clockwise order}. It is important that you enter your points
in clockwise (or counter-clockwise) order, otherwise the boundary
generation will not work.

Writing your own internal boundaries can be difficult. To make life
easier, I wrote the \textsc{Matlab}/\textsc{Octave} program
\verb+makezone.m+ to do this for us. It's in the folder \verb+m-files+
in the same repository where you found this tutorial. Run, in
\textsc{Matlab} or \textsc{Octave}

\qquad \verb+>> help makezone+

to see what input is required. The simplest way of using it is to give
it the surface points (or shallower subsurface points, if you want a completely
buried zone), the depth of the zone under the surface (let's say we
want the internal zone to end at 5 m depth), and the number of
the first point (we already have 10 points in our list, so the first
point will be 11).  Let's say we want the text output to be stored in
the file \verb+myfirstzone.txt+. To do all of this, run in
\textsc{Matlab} or \textsc{Octave}

\qquad \verb+>> makezone([-2,-2,2,2],[-2,4,4,-2],[-0.3,0.2,0,0.1],-5,11,'myfirstzone.txt')+


The resulting file \verb+myfirstzone.txt+ will contain the points
with point numbers between 11 and 18, and information about internal
planes. Add the points to your list of points in your
\verb+myfirstmesh.cfg+ file (don't forget to update the total number
of points) and replace the line saying that there are
no internal planes with the internal planes you generated.

So the points part of your file \verb+myfirstmesh.cfg+ will now be


\verb+18   (number of points)+\\
\verb+1 -10 -30 0 2   (first boundary point)+\\
\verb+2 50 -5 1 2  (second boundary point)+\\
\verb+3 40 70 -1.2 2  (etc)+\\
\verb+4 -20 60 0 2+\\
\verb++\\
\verb+5 0 0 0 1   (first electrode)+\\
\verb+6 0 1 0.5 1  (second electrode)+\\
\verb+7 0 2 0.2 1  (third electrode)+\\
\verb++\\
\verb+8 0 0 -0.01 0   (first electrode depth point)+\\
\verb+9 0 1 0.49 0  (second electrode depth point)+\\
\verb+10 0 2 0.19 0  (third electrode depth point)+\\
\verb++\\
\verb+11 -2.000000 -2.000000 -0.300000 1 +\\
\verb+12 -2.000000 4.000000 0.200000 1 +\\
\verb+13 2.000000 4.000000 0.000000 1 +\\
\verb+14 2.000000 -2.000000 0.100000 1 +\\
\verb++\\
\verb+15 -2.000000 -2.000000 -5.000000 0 +\\
\verb+16 -2.000000 4.000000 -5.000000 0 +\\
\verb+17 2.000000 4.000000 -5.000000 0 +\\
\verb+18 2.000000 -2.000000 -5.000000 0 +\\

And the internal planes part will now be

\verb+5 number of internal planes+\\
\verb+4 10+\\
\verb+11 12 16 15+\\
\verb+4 10+\\
\verb+12 13 17 16+\\
\verb+4 10+\\
\verb+13 14 18 17+\\
\verb+4 10+\\
\verb+11 14 18 15+\\
\verb+4 10+\\
\verb+15 16 17 18+\\ 
\verb++\\

Note that here we used the internal boundary number 10, which I set as
default. If you have several internal zones, then you may want to use
different internal boundary numbers (one for each zone).


We will also need to add that we have a second zone. We do that by
simply adding a line in the zones part of the file
\verb+myfirstmesh.cfg+ and set the maximum mesh volume and
conductivity for this zone. For this we need to find a point in the
external zone. In our case, the point (10/10/-10) should do. The point
(0/0/-2) is still within our internal zone. Let's set, for the
internal zone, the maximum mesh volume 1 m$^3$ and for the external zone
the maximum mesh volume $10^5$ m$^3$. We set 0.01 S/m for the
conductivity of both zones.

Hence our new ``zones'' part of the file \verb+myfirstmesh.cfg+ will
be

\verb+2  (two zones)+\\
\verb+1 0 0 -2 1 0.01  (internal zone with fine mesh)+\\
\verb+2 10 10 -10 1e5 0.01  (external zone with coarse mesh)+


After updating your file \verb+myfirstmesh.cfg+, run on the command
line

\qquad \verb+e4d+

and look at the new mesh using paraview.

You will see that your new mesh has two different zones and the mesh
of the internal zone is much finer than the mesh of the external zone.


Congratulations, you now have the knowledge needed to create
electrical resistivity tomography meshes for quite general settings.

\textbf{A note on topography:} As you noticed, all of our points had a
z-axis value. This is how you define topography. As long as your point
is either of \emph{type} 1 or 2, the mesh generator will know that it
is on the surface and will incorporate it into the topography. To
incorporate complex topography, simply add all the topography points
to your points list as \emph{type} 1 surface points (don't forget to
update the total number of points).

\textbf{Exercise:} Instead of a 4 point internal zone boundary, make a
5 point internal zone boundary. Make sure you enter the points in
clockwise orientation.

\textbf{Exercise:} Download topography points from an online database
(for example \url{http://opentopo.sdsc.edu/datasets}) and create a
mesh for these topography points.

To make your life easier I wrote \textsc{Matlab}/\textsc{Octave}
functions that can help with some of the tasks for mesh
generation. You can find them in the folder \verb+m-files+. For
example, the function \verb+elecs2points.m+ will write electrode
coordinates in the format that can directly be copy-pasted into your
e4d mesh configuration file. The function \verb+topodata2grid.m+ can
turn ASCII point-cloud data downloaded from
\url{http://opentopo.sdsc.edu/datasets} into the format that can be
copy-pasted into the e4d mesh configuration file (including
downsampling).


\section{Inversion}

In order to invert ERT data we need:

\begin{itemize}
\item A mesh that has the electrodes incorporated (see Section~\ref{meshsec})
\item A data file .srv
\item An inversion options file
\item An output options file
\end{itemize}

We learn how to do these things using a very simple data set collected on Fresno State Campus. The data \verb+LNWEN3D.ohm+ is in the folder \verb+tutdata+. The folder \verb+m-files+ contains \textsc{Matlab}/\textsc{Octave} functions to turn .ohm files into .srv files. Thomas Gunther's program DC2DInvRes, freely available from \url{resistivity.net} allows to transform most commonly used ERT data formats into .ohm (and allows for basic data quality control).

I suggest that you create a new folder for the mesh generation and inversion that we will do in this chapter.

\subsection{Step 1: Create the basic mesh file}
To do this we need to decide what area we want for the internal and
external zone. When opening the data file LNWEN3D.ohm with a standard
text editor we see that the 28 electrodes are all along the x-axis at
a 0.5 m spacing with the first electrode at the origin (0/0/0) and the
last electrode at (13.5/0/0). Let's make a mesh with external zone
between the four points (-100/-100/0), (-100/115/0), (100/115/0), and
(100/-100/0). For the internal zone we pick (-5/-5/0), (-5/5/0),
(20/5/0), and (20/-5/0). Make the internal zone 17 m deep and the
external zone 50 m deep. For the internal zone, let's make the maximum
mesh size 5 m$^3$ and for the external zone $10^7$ m$^3$.

\textbf{Task:} Create a mesh for the given external- and internal-zone
boundary points. Let's call the mesh .cfg file
\verb+myfirstinvmesh.cfg+. You will also need to create a new
\verb+e4d.inp+ file. In the \verb+myfirstinvmesh.cfg+ file, do not yet
add the electrodes. We will use the functions in the folder
\verb+m-files+ to make that part easier.

\textbf{Hint:} If the mesh generation fails, you can try making the
internal zone a bit deeper or move the boundary points around. For
example, try to create the mesh with the current settings but change
the depth of the internal zone to 10 m. The mesh generation will
(probably) fail.


\subsection{Step 2: Create the data file and add electrode positions to mesh file}

We will transform a given .ohm file (see
\url{http://resistivity.net/}) into an .srv file and use the loaded
electrode positions to add them to the mesh file.

The function \verb+OHM2E4D.m+ will do this for us. In addition, it
will return the electrode positions. Look at the help page of
\verb+OHM2E4D.m+ to understand how to use it. The provided data file
\verb+LNWEN3D.ohm+ does not have any data errors (open it with a text
editor, scroll down to the measured data: The column with the error is
missing). This means we need to estimate the error (in percent). For
this example, let's say it's 5 \%.

So we need to run in \textsc{Matlab}/\textsc{Octave}

\qquad \verb+>> electrodes=OHM2E4D('LNWEN3D.ohm','LNWEN3D',5);+

Let's also save the electrodes

\qquad \verb+>> save('electrodes.mat','electrodes');+

When you open the file \verb+LNWEN3D.srv+ in a text editor, you will
find that it looks similar to the original .ohm file, but the points
are in E4D format, the data are numbered, and there is a column for
data error.

Let's use \verb+elecs2points.m+ to create the electrode points for the mesh
file. Remember we also need to create electrode depth points to refine
the mesh around the electrodes. Let's put the depth points at 1 cm
depth. For wider electrode spacings you can set this depth to a
greater value.

Remember that our points need to be numbered and that we already have
12 points in our \verb+myfirstinvmesh.cfg+ file. So we need to set
shiftpointnr to 12.

\qquad
\verb+>> elecs2points(electrodes(:,1),electrodes(:,2),electrodes(:,3),...+\\
\hspace{1cm} \verb+'elecs.txt',12,0.01)+

creates a file \verb+elecs.txt+, from which we can copy the electrode
points and the electrode depth points and paste them into our mesh
.cfg file \verb+myfirstinvmesh.cfg+. \textbf{Remember to set the
  number of points to the new correct value}. We now have,
alltogether, 68 points.

Run

\verb+e4d+

to generate the mesh and look at it using Paraview. The resulting mesh
should have two zones. The external zone should have a very coarse
mesh, the internal zone should have a mesh that is coarse at the
boundary to the external zone and gets finer toward the electrodes. At
the electrodes, the mesh should be very fine.

\subsection{Inversion options file}

We now need to write an inversion configuration file. The E4D user
guide \url{https://e4d.pnnl.gov/Documents/E4D_User_Guide.pdf} has
more detailed descriptions on how to write inversion options
files. Here we create the simplest possible file for our two-zone situation.

I took the following inversion file pretty much directly from the E4D user
guide. Open a text file named \verb+myinvoptions.inv+ and write in it

\newpage

\verb+2                 # of constraint blocks+ \\
\verb++\\
\verb+1                 constraint zone 1     + \\            
\verb+2  1.0 1.0 1.0    structural metric 2 (wx wy wz ignored) + \\
\verb+1  10.0 .01       weight func. 1 with mean of 10 and small s.d+ \\
\verb+1 2               one link to zone 2+ \\
\verb+0.0               reference value (not used)+ \\
\verb+1.0               relative weight+ \\
\verb++\\
\verb+2                 zone # this block constrains  + \\            
\verb+2  1.0 1.0 1.0    s_met wx wy wz + \\
\verb+1  10.0 .01       fw mn sd     + \\
\verb+0                 no links, these are given already + \\
\verb+0.0               v_ref+ \\
\verb+1.0               w_rel+ \\
\verb++ \\
\verb+100 0.25 0.5      beta min_red beta_red+ \\
\verb+1.0               chi_targ+ \\
\verb+30 50             miniter maxiter+ \\
\verb+0.00001 1.0       minsig maxsig+ \\
\verb+2                 up_opt+ \\
\verb+0                 number of scaling factors+\\
\verb+1 3.0             cflag cdev+


\vspace{1cm}
I recommend studying Section 3 of the E4D user guide to understand
what these parameters mean and how changing them will affect the
inversion outcome. 

\subsection{Running the inversion}

We now need to adapt the e4d input file \verb+e4d.inp+. I recommend to
save the mesh generation version as \verb+e4d_mesh.inp+ such that we
can use it again later.

To run an inversion, e4d input file needs to look as follows:

\verb+3                      (run mode; 3 means inversion. 1 was mesh generation)+\\
\verb+myfirstinvmesh.1.node  (the name of your mesh file with .1.node appended)+\\
\verb+LNWEN3D.srv            (The data file)+\\
\verb+'average'              (How to calculate starting conductivity)+\\
\verb+myoutput.out           (Output options file; not important at this point)+\\
\verb+myinvoptions.inv       (Inversion options file)+\\
\verb+none                   (reference model)+

\newpage

To run the inversion calculation, simply running e4d will not
work. These computations are usually work-intense and e4d is
parallelized to make the calculations much faster. Therefore to
calculate the inversion we need to run e4d in parallel mode. Let's
start by using 5 processors:

\qquad \verb+mpirun -np 5 e4d+

To run it with 10 parallel processors, simply change 5 to 10 in the
above command. The number of processors you want to use is limited by
the number of processors you machine has.


\subsection{Viewing the inversion results}

During the inversion, E4D writes out the resulting conductivities for
each iteration. These are named sigma.1, sigma.2, etc., and are simple
text files with numbers in them. Each number represents the resulting
conductivity for that particular cell of the mesh. To turn the
conductivities into an .exo file, say we want to create the file
\verb+myconductivities.exo+ from the inversion result \verb+sigma.2+,
run in a command line

\qquad \verb+bx -f myfirstinvmesh.1 sigma.2 myconductivities.exo 0+

\textbf{Usually your inversion will run for more than just 2
  iterations. I recommend to take the last inversion result, for
  example sigma.23 if it ran for 23 iterations.}

\textbf{Hint:} When looking at your result in Paraview, you have to
select the variable \verb+real_conductivity+ before you click on
``Apply''. I also recommend to un-check the ``Unnamed  block ID: 2''
to only show the internal zone. Finally, once you clicked on
``Apply'', you have to select \verb+real_conductivity+ as the
parameter to display the pseudocolors.   


When viewing your inversion result in Paraview you will notice that 1)
they don't look very interesting and 2) the values are ``odd''. This
is because we display the conductivity in S/m and not the resistivity
in $\Omega$ m. To turn conductivity into resistivity, there are two
ways. Either use the \textsc{Matlab}/\textsc{Octave} program
\verb+sigma2rho.m+

\qquad \verb+>> sigma2rho('sigma.2','rho.2')+

or, if you have GMT installed (which we have on
the Student computer), run in the command line

\qquad \verb#gmt math 1 sigma.2 DIV -hi = rho.2#

and then delete the second and third row and remove the \# sybmol from
the first row. Editing the rho.2 file is only necessary when you use GMT to calculate
the resistivities.

Congratulations! You calculated your first inversion. The resulting
resistivity is symmetric because we only have profile data. To get
more interesting inversion results you can either work with your own
data, or use data from the E4D examples obtainable from \url{https://e4d.pnnl.gov/Documents/e4d_tutorial.tgz}

\textbf{Exercise:} Refine the mesh for the internal zone and rerun the
inversion. Hint: The mesh size is controlled by the maximum mesh size
parameter in your \verb+myfirstinvmesh.cfg+ file, where you define the
zones. Other hint: You will need to recreate the mesh, which means
renaming  \verb+e4d.inp+ to, for example \verb+e4d_inv.inp+ and
renaming \verb+e4d_mesh.inp+ to \verb+e4d.inp+ and then, once the mesh
is created, rename them back.

\textbf{Exercise:} Play around with the inversion options. See how
they influence the inversion result.



\end{document}


